### 2018-04-02 ident/docs/draft_1.tex
changes to methods + move table to end

	In order to establish the practical identifiability of kinetic parameters for each flux $v_i$, $i={1, 2, ..., n_r}$, we describe a computer algebra-based method. The primary use of the computer algebra system is to obtain closed-form expressions for each parameter in $\mathbf{\theta}$ for each flux $v_i$ (Figure \ref{fig:ident-flowchart}b). This is done by first selecting a combination of $p$ data sets from the $n_{E}$ available experimental data sets. The combination consisting of data from $p$ different experiments is then used to formulate a system of nonlinear algebraic equations in $\mathbb{R}^p$ for each flux $v_i$, shown in Equation (\ref{eq:nonlineq}). 
	\begin{align}\label{eq:nonlineq}
	v_{i, j} = f_j(\mathbf{x}_j,\mathbf{\theta}, \mathbf{u}_j) && \forall j=\{1, 2, ..., p\}\subset\{1, 2, ..., n_E\}
	\end{align}
	Here, $v_{i,j}$ refers to the value of the flux $v_i$ obtained from experiment $j$, $\mathbf{x}_j$ and $\mathbf{u}_j$ are the vector of metabolite and other input concentrations from each experiment $j$. Note that $\mathbf{\theta}$ is a vector in $\mathbb{R}^p$.

	Each equation in (\ref{eq:nonlineq}), indicated by the index $j$, corresponds to the kinetic rate law expression $f(x, \theta, u)$ for each $v_i$, $i={1, 2, ..., n_r}$, described in Equation (\ref{eq:kinstoich}b), written for concentrations ($\mathbf{x}_j$, $\mathbf{u}_j$) and fluxes ($v_{i,j}$) obtained from experiment $j$. Solving the system in Equation (\ref{eq:nonlineq}) results in $\mathbb{R}^p$ nonlinear expressions for parameters in $\theta$ (Equation \ref{eq:theta-eq}), where $N(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ is the numerator of $g$, and $D(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ is the denominator of $g$ (Figure \ref{fig:ident-flowchart}b). Note that $\mathbf{v}_i$, $\mathbf{x}$ and $\mathbf{u}$ are used to denote a vector of vectors of fluxes for reaction $i$ ($\mathbf{v}_i$), metabolite ($\mathbf{x}$) and input ($\mathbf{u}$) concentrations, respectively, obtained from all chosen experiments $j = {1, 2, ..., p}$, in the selected combination.
	\begin{align}\label{eq:theta-eq}
	\theta_j = g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = \frac{N_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})}{D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})}
	\end{align}
	The identifiability of parameter $\theta_j$ for flux $v_i$ can be established by determining the value of $D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ (Figure \ref{fig:ident-flowchart}b): any parameter $\theta_j$ is said to practically identifiable if $D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})\neq0$ and practically non-identifiable if $D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = 0$. Furthermore, the physical properties of the kinetic parameters can be used to distinguish between identifiable and non-identifiable parameter values by designating only parameters with a non-negative value of $g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ as identifiable (Figure \ref{fig:ident-flowchart}b). The solution obtained for $g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ in Equation (\ref{eq:theta-eq}) is unique for an identifiable $\theta_j$, and an infinite number of are possible for a non-identifiable $\theta_j$. However, if there are multiple but finite solutions for $g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$, then the corresponding parameter $\theta_j$ is locally identifiable.

	For completeness, in the following sections we provide a previously published kinetic model of a small gluconeogenic network (section \ref{sec:small-model}). In the following section we demonstrate our methodology to establish practical identifiability for one of the fluxes in this network (section \ref{sec:example}) (move to results).

@@

	In total, based on the above discussion, we can perturb four different model parameters ($acetate$, $k_1^{cat}$, $V_2^{max}$ and $V_3^{max}$) to obtain experimental data in silico. The 18 different parameter values used to generate experimental data are given in Table \ref{tab:pval}.	
	\begin{table}[!thbp]
		\caption{Table showing the perturbed values of all parameters used to generate experimental data in silico for testing practical identifiability of all fluxes in the small metabolic network.}
		\begin{center}				
			\begin{tabular}{lp{3.5cm}cccccp{2cm}}
				\hline
				Experiment Type & Perturbed Parameter (wild type value a.u.)& \multicolumn{5}{c}{Perturbed Values a.u.} & Number of Experiments\\
				\hline
				wild type &  &\multicolumn{5}{l}{} & 1\\
				acetate perturbation & $acetate$ (0.1) & 0.05 & 0.09 & 0.2 & 0.5 & 1.0 & 5\\
				$v_1$ perturbation & $k_1^{cat}$ (1) & 0.5 & 0.9 & 1.1 & 1.5 & 2.0 & 5\\				
				$v_2$ perturbation & $V_2^{max}$ (1) & 0.5 & 0.9 & 1.1 & 1.5 & 2.0 & 5\\								
				$v_3$ perturbation & $V_3^{max}$ (1) & 0.5 & 0.9 & 1.1 & 1.5 & 2.0 & 5\\
				\hline
				\multicolumn{7}{l}{Total number of experimental data sets used for practical parameter identification:} & 21\\
				\hline
			\end{tabular}
		\end{center}	
		\label{tab:pval}
	\end{table}

	The minimum number of experiments from which data is required for identifying all the parameters of a given flux is determined by dimension $\mathbb{R}^p$ of the parameter space of a chosen flux $v_i$. For instance, as demonstrated above, data from two distinct experiments is required for identifying $v_1$, as $v_1$ has two parameters ($\mathbb{R}^2$). $v_2$ is also modeled by a parameter space with dimensions of $\mathbb{R}^2$. Hence, multiple combinations of data generated from any two different experiments are used to test the identifiability of $v_1$ and $v_2$. The total number of such possible combinations is 420 (21 x 20) from the 18 different experiments shown in Table \ref{tab:pval}. Similarly, for identifying $v_3$, that is described by three parameters ($\mathbb{R}^3$), we need data from a combination of three different experiments. Based on the 21 experiments in Table \ref{tab:pval}, we have 7980 distinct combinations (21 x 20 x 19) of three experiments.

	As an example, if 90\% of all the experimental data combinations used for testing can identify a parameter $\theta_i$, then the degree of identifiability of $\theta_i$ is said to be 0.9 or 90\%. On the other hand, if only 50\% of the combinations can identify another parameter $\theta_j$, then $\theta_j$ has a degree of identifiability of 0.5 or 50\%. Furthermore, we can create a hierarchy of practically identifiable parameters using their degrees of identifiability. In the above instance of the two parameters $\theta_i$ and $\theta_j$ that have degrees of identifiability of 90\% and 50\% respectively, $\theta_i$ is classified to be more identifiable than $\theta_j$ due to its relatively higher degree of identifiability. Determining this hierarchy of identifiable parameters can help in distinguishing parameters that can be identified by any type and any combination of experiments from parameters that can be identified by only a select type and combination of experiments. Such a classification can subsequently be used to design minimal sets of experiments that can practically identify all kinetic parameters used to model a metabolic network, going from the least identifiable parameter to the most identifiable parameter.

	\subsection{Experimental design through practical parameter identification}\label{sec:experimental_design}

	Following the methodology described in section \ref{sec:ident}, and demonstrated in section \ref{sec:example} for a single flux using data from a combination of two different experiments, all distinct combinations found based on experiments described in section \ref{sec:experiments} can be tested for their ability to practically identify any of the three fluxes in the small metabolic network. This step would determine the degree of identifiability (defined in section \ref{sec:degree_of_identifiability}) of each parameter in each flux in the model, and help distinguishing experiment combinations that contribute to identifiability from combinations that do not practically identify any parameter in the model (Figure \ref{fig:ident-flowchart}b). In doing so, it is possible to obtain a minimal and informative collection of experiments that can be performed to identify as many model parameters as possible (Figure \ref{fig:ident-design}). Consequently, the set of experiments can be used to estimate the identifiable parameters in the model. This is formally explained below.

@@

	\section{Results:}
	First in section \ref{sec:closedform}, we show that the ability to establish practical identifiability of kinetic parameters in metabolic network models using our methodology, described in section \ref{sec:ident}, relies upon the nonlinearity of the kinetic rate law formulations used to describe the fluxes. We use the gluconeogenic model (Figure \ref{fig:network}) described in section \ref{sec:example} as an example. 
	Then, in section \ref{sec:data_utility} we provide a motivation for the need for experimental design, especially to identify kinetic models of metabolic networks. We do this by looking at the number of combinations that can identify the maximum number of parameters. 
	In section \ref{sec:trends} we show that the degree of identifiability of the maximum reaction rate parameter is always higher than the degree of identifiability of the corresponding enzyme binding parameter irrespective of enzyme kinetic rate law used to describe the flux. 
	In section \ref{sec:experiment_type} we discuss the ability to determine the type of experiments for parameter identification for a given flux based on the informativeness of a given type of experiment. In this section we show how the informativeness of a given type of experiment to identify a specific flux can be deduced from its contribution towards the practical identification of the parameters for a given flux. Finally, we discuss some results arising out of the use of our methodology to determine the identifiability of parameters when data with additive noise is used in section \ref{sec:noise}.

@@

	\clearpage
	\section{Figures and Figure Captions:}
	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=.6\textwidth,height=.6\textheight,keepaspectratio]{figures/figure3/ident_analysis}}
		\caption{A flow diagram showing the methodology developed to establish practical identifiability of parameters in kinetic models of metabolism. a) The steps for the construction of a kinetic model of a metabolic network. The choice of rate law formulations to describe metabolic fluxes influences the identification methodology. The identifiability of parameters for each flux can be established independently. b) The steps for practical identifiability analysis for parameters of a single flux.}\label{fig:ident-flowchart}
	\end{figure}

>>>>

	In order to establish the practical identifiability of kinetic parameters for each flux $v_i$, $i={1, 2, ..., n_r}$, we describe a computer algebra-based method. The primary use of the computer algebra system is to obtain closed-form expressions for each parameter in $\mathbf{\theta}$ for each flux $v_i$ (Figure \ref{fig:ident-flowchart}b). This is done by first selecting a combination of $p$ data sets from the $n_{E}$ available experimental data sets. The combination consisting of data from $p$ different experiments is then used to formulate a system of nonlinear algebraic equations in $\mathbb{R}^p$ for each flux $v_i$, shown in Equation (\ref{eq:nonlineq}). 
	\begin{align}\label{eq:nonlineq}
	v_{i, j} = f_j(\mathbf{x}_j,\mathbf{\theta}, \mathbf{u}_j) && \forall j=\{1, 2, ..., p\}\subset\{1, 2, ..., n_E\}
	\end{align}
	Here, $v_{i,j}$ refers to the value of the flux $v_i$ obtained from experiment $j$, $\mathbf{x}_j$ and $\mathbf{u}_j$ are the vector of metabolite and other input concentrations from each experiment $j$, and $\mathbf{\theta}$ is a vector in $\mathbb{R}^p$.

	Each equation in (\ref{eq:nonlineq}), indicated by the index $j$, corresponds to the kinetic rate law expression $f(x, \theta, u)$ for each $v_i$, $i={1, 2, ..., n_r}$, described in Equation (\ref{eq:kinstoich}b), written for concentrations ($\mathbf{x}_j$, $\mathbf{u}_j$) and fluxes ($v_{i,j}$) obtained from experiment $j$. Solving the system in Equation (\ref{eq:nonlineq}) results in $\mathbb{R}^p$ nonlinear expressions for parameters in $\theta$ (Equation \ref{eq:theta-eq}), where $N(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ is the numerator of $g$, and $D(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ is the denominator of $g$ (Figure \ref{fig:ident-flowchart}b). Note that $\mathbf{v}_i$, $\mathbf{x}$ and $\mathbf{u}$ are used to denote vector of vectors of fluxes for reaction $i$ ($\mathbf{v}_i$), metabolite ($\mathbf{x}$) and input ($\mathbf{u}$) concentrations, respectively, obtained from all chosen experiments $j = {1, 2, ..., p}$, in the selected combination.
	\begin{align}\label{eq:theta-eq}
	\theta_j = g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = \frac{N_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})}{D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})}
	\end{align}
	The identifiability of parameter $\theta_j$, $j = {1, 2, ..., p}$, for flux $v_i$ can be established by determining the value of $D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ (Figure \ref{fig:ident-flowchart}b): any parameter $\theta_j$ is said to practically identifiable if $D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})\neq0$, and practically non-identifiable if $D_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = 0$. Furthermore, the physical properties of the kinetic parameters can be used to distinguish between identifiable and non-identifiable parameter values by designating only parameters with a non-negative value of $g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ as identifiable (Figure \ref{fig:ident-flowchart}b). The solution obtained for $g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ in Equation (\ref{eq:theta-eq}) is unique for an identifiable $\theta_j$, and an infinite number of solutions are possible for a non-identifiable $\theta_j$. However, if there are multiple but finite solutions for $g_j(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$, then the corresponding parameter $\theta_j$ is locally identifiable.

	%For completeness, in the following sections we provide a previously published kinetic model of a small gluconeogenic network (section \ref{sec:small-model}).

	As an example, if 90\% of all the experimental data combinations used for testing can identify a parameter $\theta_i$, then the degree of identifiability of $\theta_i$ is said to be 0.9 or 90\%. On the other hand, if only 50\% of the combinations can identify another parameter $\theta_j$, then $\theta_j$ has a degree of identifiability of 0.5 or 50\%. Furthermore, we can create a hierarchy of practically identifiable parameters using their degrees of identifiability. In the above instance of the two parameters $\theta_i$ and $\theta_j$ that have degrees of identifiability of 90\% and 50\% respectively, $\theta_i$ is classified to be more identifiable than $\theta_j$ due to its relatively higher degree of identifiability.

	Determining this hierarchy of identifiable parameters can help in distinguishing parameters that can be identified by any type and any combination of experiments from parameters that can be identified by only a select type and combination of experiments. Such a classification can subsequently be used to design minimal sets of experiments that can practically identify all kinetic parameters used to model a metabolic network, going from the least identifiable parameter to the most identifiable parameter.

@@

	In total, based on the above discussion, we can perturb four different model parameters ($acetate$, $k_1^{cat}$, $V_2^{max}$ and $V_3^{max}$) to obtain experimental data in silico. The 21 different parameter values used to generate experimental data are given in Table \ref{tab:pval}.

	As described previously in section \ref{sec:ident} and Figure \ref{fig:ident-flowchart}b, the minimum number of experiments from which steady state data is required for identifying all the parameters of a given flux is determined by dimension $\mathbb{R}^p$ of the parameter space of a chosen flux $v_i$. For instance, the flux $v_2$ in the model described in section \ref{sec:small-model} has two parameters: $V_2^{max}$ and $K_2^{pep}$. Thus $\theta = \{V_2^{max}, K_2^{pep}\} \in \mathbb{R}^2$. So, steady state data from two distinct experiments is required for identifying $v_2$. Accordingly, multiple combinations of data generated from any two different experiments are used to test the identifiability of $v_2$. The total number of such possible combinations is 420 (21 x 20) from the 21 different experiments shown in Table \ref{tab:pval}. Assuming we also have data for the enzyme concentration $E$ in the expression for flux $v_1$ (Equation \ref{eq:flux1}), identifying the two parameters $k_1^{cat}$ and $K_1^{ac}$ ($\theta \in \mathbb{R}^2$) also requires steady state experimental data from two different experiments.

	Similarly, for identifying $v_3$ that is described by three parameters ($\theta \in \mathbb{R}^3$), we need data from a combination of three different experiments. Based on the 21 experiments in Table \ref{tab:pval}, we have 7980 distinct combinations (21 x 20 x 19) of three experiments. However, it is important to note that not all of the either 420 combinations (for $v_1$ and $v_2$), or the 7980 combinations (for $v_3$), may be suitable for parameter estimation. Hence, the actual number of data combinations that can be used for practically identifying parameters in these fluxes would be smaller than the aforementioned numbers. Moreover, among these chosen smaller number of experimental data combinations, our methodology enables us to calculate the number of experimental data combinations, and consequently, the number of experiments that might allow for parameter identification, and subsequent parameter estimation. The following section describes a procedure that can be used to design experiments within this context.

	\subsection{Experimental design through practical parameter identification}\label{sec:experimental_design}	
	Following the methodology described in section \ref{sec:ident}, and demonstrated in section \ref{sec:example} for a single flux using data from a combination of two different experiments, all distinct combinations found based on experiments described in section \ref{sec:experiments} can be tested for their ability to practically identify any of the three fluxes in the small metabolic network. This step would determine the degree of identifiability (defined in section \ref{sec:degree_of_identifiability}) of each parameter in each flux in the model, and help distinguishing experiment combinations that contribute to identifiability from combinations that do not practically identify any parameter in the model (Figure \ref{fig:ident-flowchart}b). In doing so, it is possible to obtain a minimal and informative collection of experiments that can be performed to identify as many model parameters as possible (Figure \ref{fig:ident-design}). Consequently, the set of experiments can be used to estimate the identifiable parameters in the model. This is formally explained below.

@@

	\section{Results:}
	First in section \ref{sec:closedform}, we show that the ability to establish practical identifiability of kinetic parameters in metabolic network models using our methodology, described in section \ref{sec:ident}, relies upon the nonlinearity of the kinetic rate law formulations used to describe the fluxes. We use the gluconeogenic model (Figure \ref{fig:network}) described in section \ref{sec:example} as an example.

	%In the following section we demonstrate our methodology to establish practical identifiability for one of the fluxes in this network (section \ref{sec:example}) (move to results).

	Then, in section \ref{sec:data_utility} we provide a motivation for the need for experimental design, especially to identify kinetic models of metabolic networks. We do this by looking at the number of combinations that can identify the maximum number of parameters. 
	In section \ref{sec:trends} we show that the degree of identifiability of the maximum reaction rate parameter is always higher than the degree of identifiability of the corresponding enzyme binding parameter irrespective of enzyme kinetic rate law used to describe the flux. 
	In section \ref{sec:experiment_type} we discuss the ability to determine the type of experiments for parameter identification for a given flux based on the informativeness of a given type of experiment. In this section we show how the informativeness of a given type of experiment to identify a specific flux can be deduced from its contribution towards the practical identification of the parameters for a given flux. Finally, we discuss some results arising out of the use of our methodology to determine the identifiability of parameters when data with additive noise is used in section \ref{sec:noise}.

@@

	\clearpage
	\section{Tables, Figures and Figure Captions:}
	\begin{table}[!thbp]
		\caption{Table showing the perturbed values of all parameters used to generate experimental data in silico for testing practical identifiability of all fluxes in the small metabolic network.}
		\begin{center}				
			\begin{tabular}{lp{3.5cm}cccccp{2cm}}
				\hline
				Experiment Type & Perturbed Parameter (wild type value a.u.)& \multicolumn{5}{c}{Perturbed Values a.u.} & Number of Experiments\\
				\hline
				wild type &  &\multicolumn{5}{l}{} & 1\\
				acetate perturbation & $acetate$ (0.1) & 0.05 & 0.09 & 0.2 & 0.5 & 1.0 & 5\\
				$v_1$ perturbation & $k_1^{cat}$ (1) & 0.5 & 0.9 & 1.1 & 1.5 & 2.0 & 5\\				
				$v_2$ perturbation & $V_2^{max}$ (1) & 0.5 & 0.9 & 1.1 & 1.5 & 2.0 & 5\\								
				$v_3$ perturbation & $V_3^{max}$ (1) & 0.5 & 0.9 & 1.1 & 1.5 & 2.0 & 5\\
				\hline
				\multicolumn{7}{l}{Total number of experimental data sets used for practical parameter identification:} & 21\\
				\hline
			\end{tabular}
		\end{center}	
		\label{tab:pval}
	\end{table}

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=.6\textwidth,height=.6\textheight,keepaspectratio]{figures/figure3/ident_analysis}}
		\caption{A flow diagram showing the methodology developed to establish practical identifiability of parameters in kinetic models of metabolism. a) The steps for the construction of a kinetic model of a metabolic network. The choice of rate law formulations to describe metabolic fluxes influences the identification methodology. The identifiability of parameters for each flux can be established independently. b) The steps for practical identifiability analysis for parameters of a single flux.}\label{fig:ident-flowchart}
	\end{figure}
